---
title: "Distribution Verification Demo"
author: "Jocelyn Pender"
date: "27/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(BIEN)
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
library(sf)
library(gtools)
library(raster)
library(statesRcontiguous)
library(SDMTools)
require(maps)
require(sp)
require(maptools)

source("../../src/functions/clean_gbif.R") # Stores utility functions
source("../../src/functions/clim_functions.R") # Stores utility functions
```

- explain the packages used
- explain the ecosystem of available tooling to help you clean up and verify your species distribution data
-- Talk about coordinate cleaner - no demo but at least show the results and what the imported data looks like, maps etc.
- you should always still do manual inspection - this is good practice, and all the tools in the world cannot replace this step - even BIEN recommends it
- introduce the BIEN dataset, the BIEN recommendations, and other range map resources that might exist
